---
title: "Homework 2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r initial}
library(tidyverse)
library(readxl)
```

## Problem 2: Mr. Trash Wheel

```{r mr_trash_wheel}
mr_trash_wheel = read_excel("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N533", col_names = TRUE) %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(sports_balls = as.integer(sports_balls),
         trash_wheel_type = "Mister") %>%
  relocate(trash_wheel_type, dumpster)

head(mr_trash_wheel, 5)
```

```{r prof_trash_wheel}
prof_trash_wheel = read_excel("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = "Professor Trash Wheel", n_max = 113, col_names = TRUE) %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(sports_balls = as.integer(sports_balls),
         trash_wheel_type = "Professor") %>%
  relocate(trash_wheel_type, dumpster)

head(prof_trash_wheel, 5)
```

```{r bind_data}
all_trash_wheel = bind_rows(mr_trash_wheel, prof_trash_wheel)
head(all_trash_wheel, 5)

prof_trash_weight = all_trash_wheel %>%
  filter(trash_wheel_type == "Professor") %>% 
  pull(weight_tons) %>%
  sum

mr_trash_balls = all_trash_wheel %>%
  filter(trash_wheel_type == "Mister",
         year == 2020) %>% 
  pull(sports_balls) %>%
  sum
```

The `all_trash_wheel` dataset contains 524 observations on 15 variables. Key fields in `all_trash_wheel` include `trash_wheel_type` (i.e. denoted as `Mister` or `Professor`) and `dumpster` (i.e. denoted ordinally); temporal fields, such as `date`; fields that quantify the extent of trash, such as `weight_tons` and `volume_cubic_yards`; fields that are counts of particular kinds of trash, such as `plastic_bottles`, `grocery_bags`, and `sports_balls`; and `homes_powered`, an outcome related to trash incineration. Professor Trash Wheel collected a total of `r prof_trash_weight` tons of trash; in 2020, Mr. Trash Wheel collected `r mr_trash_balls` sports balls.

## Problem 3: FiveThirtyEight

```{r pols}
pols = read_csv(file = "data/pols-month.csv") %>%
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>% 
  mutate(month = month.abb[month],
         president = ifelse(prez_dem == 1, "dem", "gop")) %>%
  select(-c(prez_dem, prez_gop, day))

head(pols, 5)
```

```{r snp}
snp = read_csv(file = "data/snp.csv") %>%
  separate(date, into = c("month", "day", "year"), sep = "/", convert = TRUE) %>% 
  mutate(month = month.abb[month],
         year = as.integer(ifelse(year < 16, year + 2000, year + 1900))) %>%
  select(-day)

head(snp, 5)
```

```{r unemployment}
unemployment = read_csv("data/unemployment.csv") %>%
  pivot_longer(Jan:Dec,
               names_to = "month",
               values_to = "rate") %>%
  rename(year = Year) %>%
  mutate(year = as.integer(year)) 
```

```{r join_data}
all = full_join(pols, snp, by = c("year", "month")) %>%
  full_join(., unemployment, by = c("year", "month"))
```

The `pols` dataset shows counts of Democrats and Republicans various holding elected offices of interest (e.g. `gov_gop`, `sen_dem`), and the president's party affiliation (i.e. `president`). The `snp` dataset consists of closing values of the Standard & Poor's index (i.e. `close`). The `unemployment` dataset reports the percentage of unemployment in the United States (i.e. `rate`). The resulting `all` dataset, a composite of the `pols`, `snp`, and `unemployment` datasets, contains 828 observations on 11 variables. The range in `years` in the `all` dataset is 1947 through 2015. Key variables in `all` include temporal fields (i.e. `year` and `month`), as well as those variables previously listed. 